### 2016-08-03 latex/input/materials.tex
added script for generating plots of GICF runs; updated paper

We obtained 904
clinical microbiome samples from a study by Dr. Patrick Gillevet 
that relates to patients who suffered from hepatic encephalopathy due 
to liver cirrhosis. Specifically, the classification formulation 
was setup to distinguish between patients suffering with a specific 
clinical phenotype or not. There were a total of 239 
patients with Encephalopathy   due to liver cirrhosis (denoted 
by the ``Encephalopathy'' class), 590 patients with liver cirrhosis 
but no Encephalopathy (denoted by ``No Encephalopathy'') 
and 75 patients who were considered as 
control and did not have either of the clinical conditions (denoted by ``Control'').
%
16S rRNA metagenomic sequence read data was 
obtained from patient stool samples. On average 
there were a total of 1,464  sequence reads of length 200-400 obtained per 
patient, with 1,323,016 total reads across all patients in our dataset.

\subsection{Evaluation Metrics}
We assess the performance of our classification pipeline 
in terms of correctness and execution time.
%
Given the imbalanced nature of class distributions,  the 
performance of  binary classifiers was measured by 
F1 score, precision and recall, along with
the standard accuracy metric. We discuss 
these standard metrics in brief.

Accuracy measures the percentage of instances 
that are classified correctly and can 
be represented by 
\begin{equation}
Accuracy = (TP + TN)/ (TP + TN + FP + FN)  \label{eqn:acc} 
\end{equation}
where TP, TN, FP and FN represents true positives, true negatives, false positives and false negatives respectively.

Accuracy as an evaluation metric can be biased if one of the classes 
(positive or negative)  has a larger number of examples than the other.   
Precision measures the percentage of positive predictions that 
were correct, 
whereas recall measures the percentage of positive 
examples that were correctly predicted (or retrieved). 
We can represent Precision and Recall by \cite{Goutte}:

\begin{equation}
Precision = TP / (TP + FP). \label{eqn:prec}
\end{equation}
\begin{equation}
Recall = TP / (TP +FN). \label{eqn:roc}
\end{equation}

The F1 score captures the trade offs between precision and recall in a
single metric and is the harmonic mean of precision and recall \cite{Goutte}, given 
by:
\begin{equation}
F1-Score = 2 * (Precision * Recall)/ (Precision + Recall). \label{eqn:f1}
\end{equation}

\subsection{Software and Hardware Details}
We used the Argo computing cluster available at George Mason University. 
%
The representation generation phase using UCLUST and Kraken and and the classification phase
were 
run on one of the  compute nodes available on the cluster. The cluster is configured with 
35 Dell C8220 Compute Nodes, each with dual 
Intel Xeon E5-2670 (2.60GHz) 8 core CPUs, with 64 GB RAM. (Total Cores – 
528 and 1056 total threads, RAM \textgreater 2TB)\cite{ORC}.

Source codes for UCLUST \cite{Edgar10} and KRAKEN \cite{Wood14} were downloaded from 
their respective websites\footnote{UCLUST: http://www.drive5.com/uclust/downloads1\_{}2\_{}22q.html     
Kraken: https://ccb.jhu.edu/software/kraken/} and compiled on the Argo platform. 
%
Kraken aligns reads to an OTU database. The standard Kraken database is 160GB in size. Due to 
computational limitations we used the custom, small sized 4GB
database  in this feasibility study \cite{Kraken}.

For the  SVM-based classification\cite{Joachims08}  we used the popular 
SVM-Light \cite{Joachims08}  source code, which is publicly 
available \footnote{http://svmlight.joachims.org/}.  The linear kernel was used
and the regularization parameters were set to their default values.

\subsection{Experimental Protocol}

For evaluating the performance of our binary phenotypic classifiers, we 
split the patient samples into a training set containing 80\% of the 
patients and the test set containing 20\% of the patient samples. Every fifth sample was placed in the test set, so the relative class sizes were roughly consistent between the training and test sets.
We also performed leave-one-out cross validation (LOOCV).

In the next section, we discuss the accuracy of our 
classifier with regards to predicting the clinical 
phenotype of a given patient using 
either the unsupervised 
clustering representation with UCLUST or 
supervised OTU representation 
with Kraken. We present results for classifiers 
distinguishing patients  in the ``Encephalopathy'' class versus
the other classes, effectively 
determining whether or not a patient had Encephalopathy 
regardless of whether they had liver cirrhosis. We also 
present pairwise one-versus-one classification results that 
compared two phenotypes within the dataset, i.e.  Encephalopathy versus
No Encephalopathy, Encephalopathy versus
Control and No Encephalopathy versus Control.

>>>>

We used data from a well-known Metagenome-Wide Association Study by Qin et al. of Type 2 Diabetes (T2D) in Chinese patients \cite{qin041012}. The total dataset used in this study contains 368 patients \cite{qin041012}, but only 218 of them were labeled and available online. Each patient file was downloaded from NCBI\footnote{http://trace.ncbi.nlm.nih.gov/Traces/study/?acc=SRP011011}
and converted to FASTQ format using the SRA toolkit\footnote{http://www.ncbi.nlm.nih.gov/Traces/sra/sra.cgi?view=software}.
The labels were found on the EBI website\footnote{https://www.ebi.ac.uk/metagenomics/projects/SRP011011}. Each patient file was several gigabytes in size, with the total dataset size being over 600 gigabytes. Qin et al. develop a simple T2D classifier using a minimum redundancy–maximum relevance (mRMR) method \cite{peng05} for feature selection and an SVM for classification, based on the R packages "sideChannelAttack" and "e10171", respectively \cite{qin041012}. They achieved an AUC-ROC of 0.81 by training a classifier on 345 of the patients and using the remaining 23 as a test set \cite{qin041012}. While we did not have access to all of the same patients that they had, we were able to replicate their method and test it on the subset of data that was publicly available. The study also called for more extensive testing of gut microbiota classifiers \cite{qin041012}.

\subsection{Software and Hardware Details}
We used the ARGO computing cluster available at George Mason University \cite{argo}. The clustering and classification phases were run on one of the compute nodes available on the cluster. The cluster is configured with 35 Dell C8220 Compute Nodes, each with dual Intel Xeon E5-2670 (2.60GHz) 8 core CPUs, with 64 GB RAM. (Total Cores 528 and 1056 total threads, RAM$>$2TB) \cite{argo}. Source codes for 
SOAPdenovo2\footnote{SOAPdenovo2: http://soap.genomics.org.cn/soapdenovo.html} \cite{luo12}, UCLUST\footnote{UCLUST: http://www.drive5.com/uclust/downloads1\_{}2\_{}22q.html}    \cite{Edgar10}, and svm-light\footnote{svm-light: http://svmlight.joachims.org/} \cite{joachims08}
their respective websites
 and compiled on the Argo platform.

### 2016-08-03 latex/input/methods.tex
added script for generating plots of GICF runs; updated paper

@@

Figure \ref{fig1} provides an overview of the 
developed computational pipeline. 
%
We obtain 16S metagenomic sequence reads from patients who 
had encephalopathy and liver cirrhosis, no encephalopathy but liver cirrhosis, and controls (with neither phenotype). The input sequence 
reads are annotated using a taxonomic profiling approach (Kraken) or 
an unsupervised binning approach (UCLUST). The output of this annotation phase 
features are generated, which are passed on the binary support vector machine 
classifier (SVMs).  We used the  svm-light implementation \cite{Joachims08}  
to perform the classification of whether or not a patient 
had a clinical phenotype. We compared the classification performance of the 
SVM classifier with the two intermediate sequence representations obtained from 
UCLUST and Kraken. We also applied feature scaling and 
selection techniques to improve the classification results.

\begin{figure*}[t]
\begin{center}
\caption{Pipeline Overview. \label{fig1}}
\includegraphics[scale=0.7]{./diagrams/concept}
\end{center}
\end{figure*}

\subsection{Unsupervised Clustering using UCLUST}
\label{uclustmethods}

In this work we evaluate the use of an unsupervised approach 
for representing our input metagenome sequence reads. Specifically, we 
  use UCLUST \cite{edgar2010search} due to its superior performance in terms 
  of run time and accuracy.
%
%
%
UCLUST \cite{edgar2010search} follows a greedy, iterative clustering approach. As a first 
step, this approach identifies exact matches of fixed length between sequence pairs known 
as seeds. These seeds are then extended by allowing for a few mismatches and/or gaps between 
the aligned pairs. Seeds scoring above a certain threshold are chosen as high segment pairs (HSPs)  and 
used for further processing. As such, the step eliminates a lot of pairwise comparisons.

Then UCLUST follows an 
incremental approach for assigning the input sequences to the different bins. 
%
%
The clustering solution is initialized as an empty list. Sequences are then 
incrementally added to the clusters existing within the list. 
%
Each unassigned input sequence is compared to the cluster representatives within the list using 
the fast indexing search technique that uses the HSPs. If a match is found with one of the 
cluster representatives, then that sequence is assigned to that particular cluster or the 
input sequence forms a new cluster. 
%
UCLUST ensures
that the cluster representatives are sequences with the largest length.

\subsection{Supervised Taxonomic Identification using Kraken}

We also evaluate the use of a search based 
approach to taxonomically label our sequence reads. We use 
Kraken for this, due to its strong performance in terms of run time 
and precision. Kraken finds $k$-mers in an input 
sequence read and queries a database of OTUs. It then associates the 
read with the lowest common ancestor (LCA) of the genomes that contains $k$-mers from the read \cite{Kraken}. In order to reduce 
the number of false positives, Kraken does not classify reads for 
which it does not find sufficient evidence of any match.

This approach is designed to perform quickly and with 
high precision. However, it also involves downloading and building 
a large database to query. It is often infeasible to download and 
build the database in its entirety, so a smaller custom  
subset of the database is available for testing 
purposes. We used this 
subset of the entire database in this study (referred to as kraken-light).

\subsection{Classification using Support Vector Machines (SVMs)}

Support Vector Machines (SVMs) \cite{vap95} are 
one of the most powerful and versatile binary classifiers used 
in myriad applications. 
%
For a  binary classification problem, the support 
vector machine (SVM) \cite{vap95} framework 
maximizes the separation between the two classes.
%

Given an input set of positive training instances and 
negative training instances, the SVM classifier learns a 
mathematical function $f$  that maximizes the separation between 
the given positive and negative training instances. The function $f$ will then be able to predict whether a new instance (considered a test instance) should have a positive or negative class label.
%
For our formulation we use SVMs to make clinical phenotype prediction for each 
patient sample 
based on their  input metagenome sequence samples represented with their 
clustering or taxa membership features.

For all the available microbiome samples,  we have a set of reads processed via UCLUST or Kraken. 
%
Each patient is associated with a vector that is used by the SVM to help build $f$. Each row in the vector, called a feature, corresponds to a certain species, and the value for that row corresponds to how closely the read from the patient matched what the species was expected to look like. The exact process of assembling a feature vector for a 
patient depends on which data representation is being used. For 
UCLUST, we use each cluster seed 
as a feature. For each input 
sequence, we use the percentage match to its cluster seed as 
the value for that feature and average the 
sequence match if multiple reads from the metagenome samples 
are assigned to the same cluster by UCLUST. 
%
For Kraken, we use the taxonomic label associated 
with the read as the feature for the SVM. The percentage match of the 
read to that label is used as the feature value. Similar 
to UCLUST, we average the similarity score if multiple reads align to the 
same taxonomic class label. 
%
Our feature selection process involves discarding Kraken's unlabeled 
reads and UCLUST's clusters with only one read (singletons).

We used cross-validation and a held out set procedure (discussed in the next 
Section) to assess the performance of SVM classification using UCLUST and Kraken
as intermediate representations.

>>>>

\begin{figure*}[t]
\centering
\includegraphics[scale=0.5]{./mil-metagenomics-pipeline.png}
\caption{This diagram illustrates the entire pipeline. Patient files with FASTQ metagenomic reads are individually assembled using SOAPdenovo2, then combined into one file and clustered with UCLUST. We extract features according to the D-BoW method, and classify the patients using the feature vectors with svm-light.} \label{pipeline}
\end{figure*}

@@

As mentioned in the Introduction, our pipeline involves a number of steps, which serve a variety of purposes. For each patient file, we assembled the sequence reads, which served the dual purpose of generating larger contigs that contain more functional biological information and reducing the dataset size by discarding reads that could not be assembled. The reads that could not be assembled were unlikely to be useful anyway, since they could not be assembled with other reads into useful contigs. The clustering step assigns the contigs to certain clusters, which represent functionally similar microbes, and thus establish classes of instances that can be used as features for the classifier. We  applied our feature extraction method, which uses the D-BoW method of measuring the closest distance from a patient's reads to each cluster center. Finally, we used a standard SVM classifier to predict patient phenotype, and used several metrics to assess its accuracy. We used the SVM's decision boundary to infer information about which clusters of instances were most or least indicative of the phenotype, discussed further in the results section. Aside from the patient labels, this process is entirely de novo, and does not consult any external databases. An illustration of this pipeline can be seen in Figure \ref{pipeline} on page \pageref{pipeline}.

\subsection{Assembly with SOAPdenovo2}

For our assembly step, we used SOAPdenovo2 \cite{luo12}, which is a de novo metagenome assembly tool. This tool was also used for assembly in the MGWAS study that our data came from. We use this method in order to compare ourselves with the MGWAS study, and because Soapdenovo2 has been proven to be one of the fastest assemblers\cite{peng12}. We tested a number of different combinations of parameters, and found that the best results came when we cut reads off after 100 base pairs (reads were 180 base pairs long originally) and used a k-mer size of 51. The patient files needed to be assembled separately, in order to avoid assembling reads from different patients together. Conversely, all contigs need to be in one file for clustering, to avoid inconsistent cluster assignments between different patients. Thus, we combined the contigs from each assembled patient file into a single for clustering.

\subsection{Clustering with UCLUST}

We used UCLUST \cite{Edgar10}, which has been shown to be among the simplest, fastest, and most effective clustering methods \cite{bonder090112, sun042711}. Since our contigs were not ordered, we used the usersort option, and we set the sequence match id to 40\%, which means that two reads needed to have 40\% of the same nucleotides to be in the same cluster. For instance, between two strings of length 100, at least 40 places in each of those strings would have to contain the same nucleotide (represented as A, T, G, or C). New contigs that did not match at least 40\% to any of the existing contigs would form the seed of a new cluster. Again, this value of 40\% was found to be the best value based on our experiments with this dataset.

\subsection{Feature Extraction}

Our feature extraction operation followed the D-BoW method. For each patient, we initialized a vector with length equal to the number of clusters. The value for each place in that patient's feature vector would then be set to the closest percentage match (0.4 to 1) of the reads from that patient to the relevant cluster seed. For instance, if there are three clusters of reads, and patient A has the cluster seed for cluster 1, no reads from cluster 2, and a read with 70\% match to cluster 3, their feature vector would be~~A = [1.0~~~0.0~~~0.7]. This is illustrated in Figure \ref{pipeline}.

\subsection{Classification and Evaluation Metrics}

We performed classification with a standard SVM classifier using the generated feature vectors; in this case, we used svm-light \cite{joachims08}. The choice of classifier is not very important for D-BoW methods \cite{amores13}.

We can assess the success of our classifier in several ways. The simplest measure, accuracy, measures the percentage of instances 
that are classified correctly, represented by 
\begin{equation}
Accuracy = (TP + TN)/ (TP + TN + FP + FN)  \label{eqn:acc} 
\end{equation}
where TP, TN, FP and FN represents true positives, true negatives, false positives and false negatives respectively.

Accuracy as an evaluation metric can be biased if one of the classes 
(positive or negative)  has a larger number of examples than the other.   
Precision measures the percentage of positive predictions that 
were correct, 
whereas recall measures the percentage of positive 
examples that were correctly predicted (or retrieved). 
We can represent Precision and Recall as:

\begin{equation}
Precision = TP / (TP + FP). \label{eqn:prec}
\end{equation}
\begin{equation}
Recall = TP / (TP +FN). \label{eqn:roc}
\end{equation}

The F1 score captures the trade-offs between precision and recall in a
single metric and is the harmonic mean of precision and recall, given by:
\begin{equation}
F1-Score = 2 * (Precision * Recall)/ (Precision + Recall). \label{eqn:f1}
\end{equation}

Finally, we also use the Area Under Curve of the Receiver Operating Characteristic (AUC-ROC), which measures the performance of the classifier as the decision boundary threshold is moved. The SVM classifier generally predicts a group label to be negative if the predicted label for that group was less than 0 and predicts a group label to be positive otherwise. The AUC-ROC measures the performance of the classifier as the threshold is varied to more or less than 0. In effect, it measures how far off incorrect predictions were from being correct. AUC-ROC plots True Positive Rate versus False Positive Rate, given by:
\begin{equation}
True Positive Rate = Recall = TP / (TP + FN). \label{eqn:roc}
\end{equation}
\begin{equation}
False Positive Rate = FP / (FP + TN). \label{eqn:prec}
\end{equation}

### 2016-08-03 latex/input/related.tex
added script for generating plots of GICF runs; updated paper

\subsection{Human Microbiome and Metagenomics}

The combination of human host cells and microbial 
cells that govern several facets of human health and pathology are referred 
to as collectively by the term \emph{human microbiome} \cite{hugenholtz2008microbiology}. 
%
Using today's sequencing technologies we have the capacity to determine 
the DNA sequences of these co-existing microbial communities. 
However, 
current  genomic  technologies 
do not provide the complete genome for each individual microbe, but 
  short, contiguous subsequences from random positions of the genome. These 
  short subsequences are referred by ``reads''. 
%
Metagenome sequence assembly is defined as the process of taking the different 
sequence reads from the various microbial organisms to produce long contiguous 
sequence of DNA for each individual organism within the community mixture. 
%
The assembly process involves identifying overlapping parts of different reads 
to order them and separate them in organism-specific larger sequences.
%

The metagenome assembly problem is known to be challenging due 
  to the similarity of genomes from the different microbes, 
differing 
  abundance, diversity, complexity and varying
  genome lengths of never-sequenced before microbes within 
  the different microbiome samples. DNA sequencing machines  
  are very high throughput and produce Terabytes of data per run and 
  also produce reads with poor quality  \cite{hugenholtz2008microbiology}. Several studies 
  have highlighted the challenges associated with the metagenome analysis and assembly problem by performing a simulation study to determine the feasibility of solving the metagenome assembly problem \cite{charuvaka2011evaluation}.

  %
  Targeted metagenomics or 16S rRNA gene sequencing provides
  a
  first step for the quick and accurate characterization of
  microbial communities. 16S sequences are marker genes, which
  exists in most microbial genomes and
  have a conserved portion
  for detection (primer development) and a variable portion that allows for
    categorization within  different  taxonomic groups \cite{petrosino2009metagenomic}.
        %
        Targeted metagenomics are also effective in detecting species with low 
        abundances. However, they may not be good at discovering 
        unique species (orphans) that have never been sequenced before. % ones that are considered to be orphan (i.e., never sequenced before).

\subsection{Microbiome Informatics}

Since the human microbiome project \footnote{http://hmpdacc.org/} and 
release of publicly available metagenomic datasets began, a host 
of methods have been developed for the analysis of 
assembled metagenomes and  input  sequence reads (before assembly)
obtained from 16S rRNA genes and whole metagenomes.  
%
%
The relationship between microbiome and human health can be characterized by first
identifying the content, abundance, and functionality of the microbes within the samples. Several
computational approaches (surveyed here \cite{:mo}) have been developed for handling the vast
amount of metagenomic data to solve two related problems: (i) clustering or binning, and (ii) taxonomy profiling methods.
%

\subsubsection{Binning Methods}

The ``binning'' problem  involves grouping input short reads  such that reads within a group are 
similar to each other. This process may lead to groups that are organism-specific and is 
unsupervised in nature, being referred by clustering in the data mining community. 
%
The binning process does not attempt to provide an automated labeling of the input reads. 
%
The clusters/bins/groups obtained from an input metagenome sample 
is referred to as  Operational Taxonomic Units (OTUs), and the number of 
OTUs gives an approximation of species diversity in a sample \cite{schloss2009introducing,schloss2005introducing,sun2009esprit}.
%
These  approaches are not constrained due to the absence of a complete coverage in taxonomic databases. Some 
environmental samples contain microbial organisms that have never been cultured in a laboratory, and thus those organisms do not exist in genomic databases.
As such,  binning of 
sequence reads has several advantages: (i)
        it can lead to an improved  metagenome assembly, (ii) it can be used 
        for computing species diversity metrics \cite{bibm2012} 
        and (iii)
        the  reduced computational
        complexity within several work-flows that analyze only
        cluster representatives, instead of individual sequences
        within a sample.

CD-HIT \cite{Li01072006}, UCLUST \cite{Edgar10}, CROP \cite{Hao01032011}, MC-MinH \cite{sdm2013a} and 
MC-LSH \cite{bibm2012}
are some of the popular 
metagenome/sequence clustering approaches used for binning. 
%
UCLUST, MC-MinH and MC-LSH  are
greedy approaches that achieve computational efficiency 
by using either hash-based indexing or matching of gap-less sequences called seeds (instead 
of expensive sequence alignment) and followup with an incremental clustering approach 
that does not involve comparing all pairs of input sequences. In Section \ref{uclustmethods} we discuss 
in detail the UCLUST \cite{Edgar10}  sequence clustering approach. We use UCLUST within 
our study, known to be a state-of-the-art metagenome clustering approach in terms of 
computational performance and accuracy of results.

\subsubsection{Taxonomy Profiling Methods}
\label{sec:tax}

The taxonomy profiling problem involves 
assigning a specific label (i.e., a phylogenetic group label) to sequence 
reads or assembled metagenome contigs~\cite{phymm}. A traditional 
approach for taxonomy profiling is to formulate  a 
classification problem (supervised) that uses marker genes for
  identification of source organism of a sequence read or fragment~\cite{Woese97}. Marker
  genes are highly conserved and provide accurate identification
  of the taxonomy class~\cite{ZongzhiLiu10012008}. These approaches 
  rely on a previously annotated 
  reference dataset like RDP~\cite{Cole2005,Cole2007,Cole2008} or GreenGenes~\cite{greengenes}. These methods
  provide
  valuable community estimates but are limited to specific 
  reads or contigs (marker genes constitute a small fraction of a metagenomic
  sequence set) and have low sensitivity due to reliance
  on an incomplete and taxon-biased reference genome database.The RDP database
  implements a naive Bayes classifier 
  that uses DNA-composition features to classify 16S RNA sequence reads into taxonomic classes as defined
  by Bergey's taxonomy~\cite{ZongzhiLiu10012008}.

  Several comparative methods have been developed for the assignment of phylogenetic class based on principles of homology. Such methods  align
  reads or contigs using BLAST~\cite{altschul90a}, and assign taxonomy based on the best match with a reference database~\cite{Tringe04222005}. MEGAN~\cite{megan} and MARTA~\cite{marta} are
  metagenomic analysis and visualization programs that  make the assignment based on multiple BLAST hits and optimized parameters.  The MG-RAST~\cite{mgrast}  web server
  provides taxonomic and functional annotation by comparative searches performed across multiple reference databases. GAAS~\cite{gaas} is a novel BLAST-based tool
  that includes genome length normalization along with a similarity weighting for multiple BLAST hits to provide improved estimates.

  Composition-based methods have also been developed. They extract key sequence features such as GC composition and $k$-mer frequencies and build supervised
  classification models using those features. PhyloPythia~\cite{McHardy2007} uses a support vector machine framework~\cite{vap95} to classify long reads into taxonomic groups
  using a $k$-mer based kernel function. TETRA~\cite{tetra} correlates the $k$-mer pattern feature to different taxonomic groups. Kraken, which is used in this paper, searches a taxonomic database for the lowest common ancestor (LCA) of the genomes that contain $k$-mers from a sequence read. Phymm~\cite{phymm} trains an interpolated Markov model
  to characterize variable length subsequences  specific to different taxonomic subgroups. In combination with BLAST, Phymm shows improved classification
  accuracy for short reads of 100 base pair (bp) length. Such Markovian models have been very successful in gene finding algorithms like Glimmer~\cite{delc98nucl}.

>>>>

\subsection{Multiple Instance Learning}

The Multiple Instance Learning (MIL) problem was first described by Dietterich in the context of drug activity prediction \cite{dietterich97}. In this problem, we want to figure out which of a number of different molecules will bind to a target "binding site". Each molecule can assume a number of different 3-dimensional shapes, so even for molecules that are known to bind to the binding site, it is not necessarily known which formation of the molecule succeeds in binding. If even one shape of a given molecule binds to the binding site, it is considered a "good" molecule \cite{dietterich97}. Thus, in the original formation of MIL, a bag is classified as positive if one or more instances within it is positive, while a negative bag contains all negative instances. In the original paper, Dietterich develops a solution based on axis-parallel rectangles to solve this problem, and the MIL approach was shown to be significantly more effective than a standard supervised learning approach \cite{dietterich97}. In the late 1990s and early 2000s, a number of different approaches were developed for the original MIL problem, such as Diverse Density (DD) \cite{perez98}, EM-DD \cite{zhang01}, MI-SVM \cite{andrews02}, and MILES \cite{wang06}. A recent review of MIL by Amores created a taxonomy of these various methods and compared their effectiveness for classification \cite{amores13}.
%

However, since 2010, there has been increased interest in different formulations of the MIL problem. For instance, the problem of "key instance detection" \cite{zhou12} revolves around finding the instances that contribute the most to bag labels. A recent study focused on a formulation of the MIL problem in which bags with negative labels can contain some "positive" instances, and developed a general cost function for determining individual instance labels from group labels \cite{kotzias15}. This is significant in metagenomics because, while some diseases are caused by a single pathogen, many arise from a combination of many factors, and even patients that are "healthy" may contain small amounts of pathogens that are normally associated with disease. 
%

However, this method treats a group label simply as an aggregation of instance labels. Amores, in his taxonomy, called this type of method "Instance-based", and found this paradigm to be ineffective \cite{amores13}. Amores identifies two other paradigms, Bag-Space and Embedded-Space. An example of the latter are Bag of Words (BoW) methods, which involve the following three-step process: (i) Cluster the instances to create classes of instances; (ii) for each bag, map the clusters of instances in that bag to a feature vector; and (iii) use a standard classifier that uses the feature vectors to predict group labels \cite{amores13}. Amores found the Distance-based Bag of Words (D-BoW) method to be the second most effective of all tested methods, and the most effective one that was also time-efficient \cite{amores13}. The distinguishing feature in D-BoW methods is that the values for the feature vector represent the instance that has the smallest distance to the cluster center. Despite these recent developments in MIL, we have not found any literature that specifically applies MIL to classifying patient phenotype based on metagenomic data.

\subsection{Assembly}

The \emph{assembly} problem involves combining overlapping short reads into longer sequences called \emph{contigs}.

\subsection{Clustering}

The \emph{clustering} problem in this context  involves grouping input short reads  such that reads within a group are 
similar to each other. This process may lead to groups that are organism-specific and is 
unsupervised in nature. 
%
The clustering process does not attempt to provide an automated labelling of the input reads. 
%
The clusters/groups obtained from an input metagenome sample 
are referred to as  Operational Taxonomic Units (OTUs), and the number of 
OTUs gives an approximation of species diversity in a sample \cite{schloss2009introducing,schloss2005introducing,sun2009esprit}.
%
These  approaches are not constrained due to the absence of a complete coverage in taxonomic databases. Some 
environmental samples contain microbial organisms that have never been cultured in a laboratory, and thus those organisms do not exist in genomic databases.
As such,  clustering of 
sequence reads has several advantages: (i)
        it can lead to an improved  metagenome assembly, (ii) it can be used 
        for computing species diversity metrics \cite{bibm2012} 
        and (iii)
        the  reduced computational
        complexity within several work-flows that analyze only
        cluster representatives, instead of individual sequences
        within a sample.

CD-HIT \cite{Li01072006}, UCLUST \cite{Edgar10}, CROP \cite{Hao01032011}, MC-MinH \cite{sdm2013a} and 
MC-LSH \cite{bibm2012}
are some of the popular 
metagenome sequence clustering approaches used for binning. 
%
UCLUST, MC-MinH and MC-LSH  are
greedy approaches that achieve computational efficiency 
by using either hash-based indexing or matching of gap-less sequences called seeds (instead 
of expensive sequence alignment) and followup with an incremental clustering approach 
that does not involve comparing all pairs of input sequences. We use UCLUST within our study, which is one of the most widely used and cited metagenome clustering methods and has been shown to be amongst the most effective in terms of speed and accuracy in benchmarking studies \cite{bonder090112, sun042711}. 
%

UCLUST \cite{Edgar10} follows a greedy, iterative clustering approach. As a first 
step, this approach identifies exact matches of fixed length between sequence pairs known 
as seeds. These seeds are then extended by allowing for a few mismatches and/or gaps between 
the aligned pairs. Seeds scoring above a certain threshold are chosen as high segment pairs (HSPs)  and used for further processing. As such, the step eliminates a lot of pairwise comparisons. Then UCLUST follows an incremental approach for assigning the input sequences to the different bins. The clustering solution is initialized as an empty list. Sequences are then incrementally added to the clusters existing within the list. Each unassigned input sequence is compared to the cluster representatives within the list using the fast indexing search technique that uses the HSPs. If a match is found with one of the cluster representatives, then that sequence is assigned to that particular cluster; otherwise, the input sequence forms a new cluster. Matches are computed by comparing two reads, character by character, and dividing the number of characters at the same position in both reads that match by the length of the longer of two reads. UCLUST ensures that the cluster representatives are sequences with the largest length.
